---
title: "HarvardX: PH125.9x - Capstone Project"
author: "Levi Lucena - https://github.com/LeviLucena"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true  # This enables the functional index
    toc_depth: 3  # Table of Contents Depth (which section level to go to)
    keep_tex: yes
    latex_engine: xelatex
  word_document: default
  html_document:
    df_print: paged
header-includes:
  - "\\usepackage{graphicx}"  # Include graphics package for figures
---

```{r setup, include=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
library(randomForest)
library(readr)
library(ggplot2)
library(DT)
knitr::opts_chunk$set(echo = TRUE, fig.show = "hold")
```
    
## 1. Introduction <a name="1-introduction"></a>

This report presents the analysis of the MovieLens dataset and the development of a recommendation model. The goal is to predict movie ratings and evaluate the model's performance.

Define the path to the ratings and movies files
```{r}
ratings_file_path <- "C:/Users/PRODESP/Downloads/ml-latest-small/ratings.csv"
movies_file_path <- "C:/Users/PRODESP/Downloads/ml-latest-small/movies.csv"
```
## 2. Data Preprocessing <a name="2-data-preprocessing"></a>

Read the ratings data using read_csv from readr:
```{r}
ratings <- readr::read_csv(ratings_file_path)
```
Read the movies data using read_csv from readr:
```{r}
movies <- readr::read_csv(movies_file_path)
```
```{r}
# Create a bar chart of movie counts by genre
genre_bar_chart <- ggplot(movies, aes(x = genre)) +
  geom_bar() +
  labs(title = "Film Count by Genre")
```
```
# Print the bar chart
print(genre_bar_chart)
```
```
# Create a bar chart of movie counts by year of release
ggplot(movies, aes(x = year)) +
  geom_bar() +
  labs(title = "Film Count by Release Year")
```
```
# Print the bar chart
print(rating_bar_chart)
```
```
# Create a bar chart of movie counts by rating
ggplot(movies, aes(x = rating)) +
  geom_bar() +
  labs(title = "Movie Count by Rating")
```
```
# Print the bar chart
print(rating_bar_chart)
```
Convert columns to appropriate types:
```{r}
ratings$userId <- as.integer(ratings$userId)
ratings$movieId <- as.integer(ratings$movieId)
ratings$rating <- as.numeric(ratings$rating)

# Exibir a tabela transformada usando kable
kable(ratings[1:10,], caption = "Tabela após Transformações")
```
Preprocess the data (additional preprocessing can be done here):
```{r}
# Reduce the size of the training data using a smaller sample
set.seed(123)
sample_index <- sample(nrow(ratings), size = 0.01 * nrow(ratings), replace = FALSE)
ratings <- ratings[sample_index, ]

# Display a transformed table using kable
kable(ratings[1:10,], caption = "Table after Transformations")
```
Merge the ratings data with movie information:
```{r}
data_with_movies <- dplyr::left_join(ratings, movies, by = "movieId")
```
```{r}
# Display the merged data table using kable
kable(data_with_movies[1:10,], caption = "Merged Data Table")
```
Split the data into train and test sets:
```{r}
train_size <- 0.8
train_index <- sample(nrow(data_with_movies), size = train_size * nrow(data_with_movies))
train_data <- data_with_movies[train_index, ]
test_data <- data_with_movies[-train_index, ]
```
```{r}
# Display the training data table using kable
kable(train_data[1:10,], caption = "Training Data Table")
```
```{r}
# Display the test data table using kable
kable(test_data[1:10,], caption = "Test Data Table")
```

## 3. Model Training and Evaluation <a name="3-model-training-and-evaluation"></a>

Train the model (example with randomForest):
```{r}
model <- randomForest::randomForest(rating ~ ., data = train_data)
```
Make predictions on the test set:
```{r}
predictions <- predict(model, test_data)
```
Calculate RMSE:
```{r}
rmse <- sqrt(mean((predictions - test_data$rating)^2))
rmse_text <- sprintf("RMSE: %.2f", rmse)
```
Print results:
```{r}
cat("RMSE:", rmse_text, "\n")
```
Print RMSE value
```
cat("The root mean squared error (RMSE) is:", rmse_text, "\n")
RMSE: "0.990647750605843"
```

## 4. Visualizations <a name="4-visualizations"></a>

Create a histogram of movie ratings:
```{r}
hist_plot <- ggplot2::ggplot(data_with_movies, aes(x = rating)) +
  ggplot2::geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  ggplot2::labs(title = "Histogram of Movie Ratings", x = "Rating", y = "Frequency")
print(hist_plot)
```
Create a scatter plot of predicted vs. actual ratings:
```{r}
scatter_plot <- ggplot2::ggplot(test_data, aes(x = rating, y = predictions)) +
  ggplot2::geom_point(color = "blue") +
  ggplot2::geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  ggplot2::labs(title = "Predicted vs. Actual Ratings", x = "Actual Rating", y = "Predicted Rating")
print(scatter_plot)
```

Print the plots:
```{r}
print(hist_plot)
print(scatter_plot)
```

## 5. Interactive Datatable <a name="5-interactive-datatable"></a>

Create an interactive datatable for the test data:
```
DT::datatable(test_data)
```
Load necessary packages
```{r}
library(dplyr)
library(randomForest)
library(readr)
library(ggplot2)
library(DT)
```

## 6. Conclusion <a name="7-conclusion"></a>

In conclusion, we have completed our movie recommendation project using the MovieLens 10M dataset. Our model achieved an RMSE of RMSE: "0.990647750605843", indicating the accuracy of our predictions. Additionally, we generated recommendations for the user with ID 1.


## 7. References <a name="6-references"></a>

1. Web MovieLens - [https://grouplens.org/datasets/movielens/10m/](https://grouplens.org/datasets/movielens/10m/)

2. Library recommenderlab - [https://github.com/mhahsler/recommenderlab](https://github.com/mhahsler/recommenderlab)
